Experiences with Planning for Natural Language Generation

### 2008-06-22 trunk/Papers/experiences/domains.tex
added PDDL actions; changes to Planning in NLG section (GIVE subsection)

@@

The object of the GIVE Challenge (``Generating Instructions in Virtual
Environments''; Koller et al.\
\citeyear{alexander07:_shared_task_propos}) is to build an NLG system
which is able to give natural-language instructions which will guide a
human user in performing some task in a virtual environment.  From an
NLG perspective, GIVE makes for an interesting challenge because it is
a theory-neutral task that exercises all components of an NLG system,
and emphasizes the study of communication in a (simulated) physical
environment.  It also has the advantage that the user and the NLG
system can be physically in different places, as long as the 3D client
and the NLG system are connected over a network.  This makes it
possible to evaluate GIVE NLG systems on a large scale over the
Internet.  The first GIVE evaluation will take place in late 2008;
currently eight research teams from five countries are working on
developing systems to participate in the challenge.

\begin{figure}
\centering
\includegraphics[width=1 \columnwidth]{give_world_no_expl}
\caption{Map of the GIVE development world.}
  \label{fig:give-development-world}
\end{figure}

A map of an example world is shown in
Fig.~\ref{fig:give-development-world}.  In this world, the user's task
is to pick up a trophy in the top left room.  The trophy is hidden in
a safe behind a picture, and the user must therefore first move the
picture away and open the safe by pushing a sequence of buttons (the
small square boxes on the walls) in the correct order.  In order to
get to all these buttons, they must also open the lower door and
deactivate the alarm tile by pushing further buttons.  To simplify
both the planning and the NLG task, the world is discretized into
tiles of equal size; the user can turn by 90 degree steps in either
direction, and can move from the center of one tile to the center of
the next.  Some actions that the user can take are shown in
Fig.~\ref{fig:give-planning}.  The first few steps of one shortest
plan to achieve the goal are:
\begin{enumerate}
\item $\mathsf{turn-left}(\mathsf{north}, \mathsf{west})$
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
\item $\mathsf{manipulate-b1-off-on}(\mathsf{pos\_5\_2})$
\item $\mathsf{turn-right}(\mathsf{west}, \mathsf{north})$
\end{enumerate}

\begin{figure}
  \centering
  xxxxx
  \caption{Excerpt from the GIVE planning problem.}
  \label{fig:give-planning}
\end{figure}

This shortest plan consists of 108 action instances, most of which are
``move'' actions.  The bulk of the GIVE problem is therefore a variant
of the Gridworld problem, which also involves finding a route through
a world with discrete positions -- but with the need to press buttons
in the right order and with somewhat more complicated room shapes.  In
this example, the NLG system could generate an instruction sequence
starting with ``turn left; walk forward; press the button; turn right;
walk forward; walk forward; turn right; walk forward; walk forward;
turn left; walk forward''.  However, this would be clumsy and boring,
and it is to be expected that frustrated users will quickly cancel an
interaction with such a system, resulting in a bad evaluation score.
Having just pressed the button, it would be much better to say ``turn
around and walk through the door''.  This means that the mapping from
action instances to natural language instructions is not trivial.
Conversely, it may also be necessary to express a single planning step
with several instructions.  Having just entered the top left room, it
may be easier for the user to understand the instruction ``walk to the
center of the room; turn right; now press the green button in front of
you'' than the instruction ``press the green button on the wall to
your right''.  To plan the referring expression ``the green button in
front of you'' at a time when the button is not yet in front of the
user, the NLG system must keep track of the hypothetical changes in
what objects are visible to the user.  Thus the acts of referring and
instructing, and the modules for planning and NLG, must be integrated
tightly.

\todo{equivalent plans; plan recognition; real-time requirements}

>>>>

\begin{figure}[t]
\centering
{\small%
\begin{verbatim}
(:action add-sleeps
   :parameters (?u - node
                ?xself - individual
                ?xsubj - individual)
   :precondition (and (subst S ?u)
      (referent ?u ?xself)
      (sleep ?xself ?xsubj))
   :effect (and (not (subst S ?u))
      (expressed sleep ?xself ?xsubj)
      (subst NP (subj ?u))
      (referent (subj ?u) ?xsubj)
      (forall (?y - individual)
         (when (not (= ?y ?xself))
            (distractor (subj ?u) ?y)))))

(:action add-rabbit
   :parameters (?u - node
                ?xself - individual)
   :precondition (and (subst NP ?u)
      (referent ?u ?xself)
      (rabbit ?xself))
   :effect (and (not (subst NP ?u))
      (canadjoin N ?u)
      (forall (?y - individual)
         (when (not (rabbit ?y))
           (not (distractor ?u ?y))))))

(:action add-white
   :parameters (?u - node
                ?xself - individual)
   :precondition (and (canadjoin N ?u)
      (referent ?u ?xself)
      (rabbit ?xself))
   :effect (forall (?y - individual)
      (when (not (white ?y))
         (not (distractor ?u ?y)))))
\end{verbatim}}%
\caption{PDDL actions for generating the sentence ``The white rabbit
sleeps''.}
\label{fig:pddl-crisp}
\end{figure}

@@

The object of the GIVE Challenge (``Generating Instructions in Virtual
Environments''; Koller et al.\ \citeyear{alexander07:_shared_task_propos})
is to build an NLG system which is able to produce natural-language
instructions which will guide a human user in performing some task in a
virtual environment. From an NLG perspective, GIVE makes for an
interesting challenge because it is a theory-neutral task that exercises
all components of an NLG system, and emphasizes the study of communication
in a (simulated) physical environment. It also has the advantage that the
user and the NLG system can be physically in different places, as long as
the 3D client and the NLG system are connected over a network. This makes
it possible to evaluate GIVE NLG systems on a large scale over the
Internet. The first GIVE evaluation will take place in late 2008;
currently eight research teams from five countries are working on
developing systems to participate in the challenge.\footnote{See the GIVE
 website for more details about this project:
 \url{homepages.inf.ed.ac.uk/v1akolle/proj/give/}.}

\begin{figure}
\centering
%\includegraphics[width=1 \columnwidth]{give_world_no_expl}
\includegraphics[width=1 \columnwidth]{give_world_2}
\caption{Map of the GIVE development world.}
  \label{fig:give-development-world}
\end{figure}

A map of an example GIVE world is shown in
Fig.~\ref{fig:give-development-world}.  In this world, the user's task is
to pick up a trophy in the top left room.  The trophy is hidden in a safe
behind a picture, so the user must first move the picture out of the way
and open the safe by pushing a sequence of buttons (the small square boxes
on the walls) in the correct order.  In order to get to all these buttons,
the user must also open the lower door and deactivate the alarm tile by
pushing further buttons.  To simplify both the planning and the NLG task,
the world is discretized into tiles of equal size; the user can turn by 90
degree steps in either direction, and can move from the centre of one tile
to the centre of the next. Some of the actions the user can take are shown
in Fig.~\ref{fig:give-planning}, in PDDL syntax.

\begin{figure}[t]
{\small%
\begin{verbatim}
(:action move
   :parameters (?from - position
                ?to - position
                ?ori - orientation)
   :precondition (and (player-pos ?from) 
      (adjacent ?from ?to ?ori) 
      (player-orient ?ori)
      (not-blocked ?from ?to)
      (not-alarmed ?to))
   :effect (and (not (player-pos ?from))
      (player-pos ?to)))

(:action turn-left
   :parameters (?ori - orientation
                ?newOri - orientation)
   :precondition (and (player-orient ?ori)
      (next-orient-left ?ori ?newOri))
   :effect (and (not (player-orient ?ori))
      (player-orient ?newOri)))

(:action turn-right
   :parameters (?ori - orientation
                ?newOri - orientation)
   :precondition (and (player-orient ?ori)
      (next-orient-right ?ori ?newOri))
   :effect (and (not (player-orient ?ori))
      (player-orient ?newOri)))

(:action manipulate-b1-off-on
   :parameters (?pos - position)
   :precondition (and (state b1 off)
      (player-pos ?pos)
      (position b1 ?pos))
   :effect (and (not (state b1 off))
      (state b1 on)
      (not (state d1 closed))
      (state d1 open) 
      (not (blocked pos_6_5 pos_6_4))
      (not (blocked pos_6_4 pos_6_5))))
\end{verbatim}}%
\caption{PDDL actions for the GIVE domain.}
\label{fig:give-planning}
\end{figure}

The plans generated in the GIVE domain are often nontrivial. For instance,
in the example above the shortest plan consists of 108 action steps; the
first few steps are as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north}, \mathsf{west})$
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$
\end{enumerate}
%
(Most of the actions in this plan are ``move'' actions.) The bulk of the
GIVE problem can be considered a variant of the Gridworld problem, which
also involves finding a route through a world with discrete positions---but
with the additional need to press buttons in the right order, reason about
many more objects in the world, and navigate somewhat more complicated room
shapes.

The task of generating natural language instructions in this domain is also
more complicated than just a straightforward translation of the action
steps in a plan. For instance, the NLG system could generate an instruction
sequence starting with ``turn left; walk forward; press the button; turn
right; walk forward; walk forward; turn right; walk forward; walk forward;
turn left; walk forward''.  However, this would be clumsy and boring, and
it is to be expected that frustrated users will quickly cancel an
interaction with such a system, resulting in a bad evaluation score.
Having just pressed the button, it would be much better to say ``turn
around and walk through the door''.  This means that the mapping from
action instances to natural language instructions is not trivial.
Conversely, it may also be necessary to express a single planning step with
several instructions.  Having just entered the top left room, it may be
easier for the user to understand the instruction ``walk to the centre of
the room; turn right; now press the green button in front of you'' than the
instruction ``press the green button on the wall to your right''.  To plan
the referring expression ``the green button in front of you'' at a time
when the button is not yet in front of the user, the NLG system must keep
track of the hypothetical changes in what objects are visible to the user.
Thus the acts of referring and instructing, and the modules for planning
and NLG, must be tightly integrated.

Plan execution monitoring also plays an important role in the GIVE problem.
At a high level, the system needs to monitor a user's actions and compare
them against the generated instruction set to determine if the user has
correctly followed directions or not. In the case of the latter, new
instructions may have to be generated. In practice, the situation
can be quite complicated since the mental state of the user is not known
and so the system must observe the user's actions in real time. For
example, a user directed to ``turn around and walk through the door'' may
not necessarily perform these actions to the letter, i.e., immediately
turning 180 degrees and proceeding directly to the door. Instead, the user
might take a roundabout route through the room, eventually exiting out the
door. Although the user's actions do not match the generated instructions
exactly, they meet the intended goal. The system must be able to identify
such ``equivalent'' plans and not immediately generate new instructions as
soon as the user's actions have gone off course. Furthermore, a user can
communicate certain intentions to the system, both through action and
inaction. For instance, the system may correctly infer that a user has
failed to follow instructions if the user exits a room when given a
directive to ``walk to the centre of the room''. However, the system should
also be able to make a similar conclusion if a user simply does nothing
when given the instruction. Thus, such real-time plan monitoring is
essential in the GIVE domain, and also affects other modules in the system.

### 2008-06-22 trunk/Papers/experiences/experiences.tex
a bit about GIVE
changes to author block and introduction
added PDDL actions; changes to Planning in NLG section (GIVE subsection)

\usepackage{aaai} 
\usepackage{times} 
\usepackage{helvet} 
\usepackage{courier} 
\title{Experiences with planning for natural language generation}
\author{Alexander Koller \and Ron Petrick \\
School of Informatics \\
University of Edinburgh}

>>>>

\usepackage{aaai} 
\usepackage{times} 
\usepackage{helvet} 
\usepackage{courier}

\usepackage{graphicx}
\usepackage{url}

\newcommand{\todo}[1]{\textbf{(#1)}}

\title{Experiences with Planning for Natural Language Generation}
\author{Alexander Koller \and Ronald Petrick \\
School of Informatics \\
University of Edinburgh \\
Edinburgh \ EH8 9AB, Scotland, UK \\
\texttt{\{a.koller,r.petrick\}@ed.ac.uk}}

\nocopyright

### 2008-06-22 trunk/Papers/experiences/experiments.tex
minor edits
added PDDL actions; changes to Planning in NLG section (GIVE subsection)

As a first impression, it is a good idea to use planning in both of
these domains.  In the GIVE domain, SGPLAN 5.2.2 computes a domain
plan from the initial state to the goal in 0.3 seconds, which is fast
enough for the application.\footnote{All runtimes reported here were
  measured on a Pentium 4 CPU at 3 GHz.  Java programs were allowed to
  ``warm up'', i.e.\ the planner was run five times in a row and the
  first four measurements discarded to ensure that the JVM had
  just-in-time compiled all relevant bytecode.}  In the sentence
generation domain, FF dramatically outperforms the best previously
known algorithm for the same problem (a reimplementation of
\cite{Stone2003a}), although the latter is a greedy search algorithm
with a heuristic that is hand-tailored to the domain.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{pic-runtime-modifiers-with-sgplan}
  \caption{Runtimes in the sentence generation domain.  The horizontal
  axis represents parameters $(m,n)$ from $(1,1)$ to $(10,10)$ in
  lexicographical order.}
  \label{fig:runtimes-crisp}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4 \textwidth]{pic-buttons}
  \caption{Minimal GIVE world.}
  \label{fig:give-minimal}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{pic-runtime-buttons}
  \caption{Runtimes of FF and SGPLAN on the minimal GIVE worlds for $h=20$. The horizontal axis is $n$, ranging from 1 to 40.}
  \label{fig:give-runtime-minimal}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{pic-empty-buttons}
  \caption{GIVE world with junk positions.}
  \label{fig:give-junk}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{pic-runtime-empty-world}
  \caption{Runtimes of FF and SGPLAN on the GIVE worlds with junk
    positions for $h=20$ and $n=5$. The horizontal axis is $w$,
    ranging from 1 to 70.}
  \label{fig:give-runtime-junk}
\end{figure}

>>>>

As a first impression, it is a good idea to use planning in both of
these domains.  In the GIVE domain, SGPLAN 5.2.2 computes a domain
plan from the initial state to the goal in 0.3 seconds, which is fast
enough for the application.\footnote{All reported runtimes were
  measured on a Pentium 4 CPU running at 3 GHz. Java programs were allowed
  to ``warm up'', i.e.\ the planner was run five times in a row and the
  first four measurements discarded to ensure that the JVM had just-in-time
  compiled all relevant bytecode.}  In the sentence
generation domain, FF dramatically outperforms the best previously known
algorithm for the same problem (a reimplementation of \cite{Stone2003a}),
although the latter is a greedy search algorithm with a heuristic that is
hand-tailored to the domain.

@@

\begin{figure}
  \centering
  \includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
  \caption{Runtimes in the sentence generation domain.  The horizontal
  axis represents parameters $(m,n)$ from $(1,1)$ to $(10,10)$ in
  lexicographical order.}
  \label{fig:runtimes-crisp}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=1\columnwidth]{pic-buttons}
  \caption{Minimal GIVE world.}
  \label{fig:give-minimal}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=1\columnwidth]{pic-runtime-buttons}
  \caption{Runtimes of FF and SGPLAN on the minimal GIVE worlds for $h=20$. The horizontal axis is $n$, ranging from 1 to 40.}
  \label{fig:give-runtime-minimal}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=1\columnwidth]{pic-empty-buttons}
  \caption{GIVE world with junk positions.}
  \label{fig:give-junk}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
  \caption{Runtimes of FF and SGPLAN on the GIVE worlds with junk
    positions for $h=20$ and $n=5$. The horizontal axis is $w$,
    ranging from 1 to 70.}
  \label{fig:give-runtime-junk}
\end{figure}
